---
title: "**Cat Anatomy Report**"
author: |
  | **Andrew Wang** 
  North Carolina School of Science and Mathematics
date: "**August 30, 2020**"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clean up and setup
rm(list=ls()) # clean up any old stuff in R
setwd("C:/Users/hyper/OneDrive/Desktop/Desktop Folders/Programming/R") # go to this folder
#load up myfunctions.R
source("C:/Users/hyper/OneDrive/Desktop/Desktop Folders/Programming/R/myfunctions.R")
library(MASS)
```

<center> **Abstract** </center>

This document provides a report on data related to cat anatomy, specifically heart weight in grams, body weight in kilograms, and gender

### **Model Setup**

In this section, we established the basic parameters of the model, including cleaning up old data, setting up the working directory, and loading a separate file of useful functions.

```{r mSetup, echo=TRUE}
rm(list=ls())
setwd("C:/Users/hyper/OneDrive/Desktop/Desktop Folders/Programming/R")
source("./myfunctions.R")
```

### **Loading Libraries**

For this study, we are using the MASS package (Modern Applied Statistics with S, from Venables and Ripley). In this package, there are a number of datasets ready for analysis, including the cats dataset.

```{r loadLibraries, echo=TRUE}
# load data from the MASS library
library(MASS)
```

```{r loadLibrariesWarning, echo=TRUE}
## Warning: package 'MASS' was built under R version 3.5.2
```

### **Data Demographics**

In this section, we look at the basic structure of the data: the head of the data file, which shows the names of the columns and a representative sample of the first several rows of data; the dimension of the data (number of rows and columns); and a descriptive statistical summary of the dataset.

``` {r dataDemographics, echo-TRUE}
head(cats)
dim(cats)
summary(cats)
```

#### **Nicer looking summary table**

We can use tools such as knitr/kable to display data tables in a nicer looking format. In this section, we are displaying the summary descriptive statistics table using these tools.

``` {r summaryTable, echo=TRUE}
knitr::kable(
summary(cats),
caption = "Table 1: Summary table of cat anatomy data"
)
```

### **Data munging: rename columns, data in columns**

“Data munging” is a computer science term that basically means cleaning up the data, usually in fairly destructive ways (not always a bad thing). In this section, we are changing the existing column names (sex, bwt,hwt) to something more readable: Gender, BodyWeight and HeartWeight. Having longer column names has its problems (more typing), but does make your dataset more human-friendly! Likewise, we are changing gender from F/M to Female/Male.
``` {r dataMunging, echo=TRUE}
colnames(cats) <- c("Gender", "BodyWeight", "HeartWeight")
names(cats)
cats$Gender <- factor(cats$Gender, levels = c("F", "M"), labels = c("Female", "Male"))
head(cats)
```

### **Graphical Representations**
In exploratory data analysis (EDA), it is useful to create basic plots and graphs to get a sense of the behavior of the data. In this section, we create three types of plots: histograms, pairwise correlation plots, and a simple scatter plot, fitted with a regression line.
#### **Histograms**
It is oftentimes useful to see histograms of the numerical data, in this case the two weights. It should be noticed that the data does not follow a normal distribution.
``` {r histograms, echo=TRUE}
hist(cats$BodyWeight, main="Histogram of Body weight (kg)")
hist(cats$HeartWeight, main="Histogram of Heart weight (g)")
```
Pairwise plot of relevant data 

We now use one of the builtin functions from myfunctions.R to create a pairwise plot. This shows histograms of all of the data, a scatter plot of each pair of data, and a correlation coefficient between each pair.

``` {r histograms2, echo=TRUE}
pairs(cats,upper.panel=panel.cor,diag.panel=panel.hist)
```

#### **Plot of heart weight as a function of body weight (with linear regression line)**
Finally, we run a scatter plot of heart weight as a function of body weight. Notice that the data is fairly linear, supported by plotting a linear regression line on top of the raw data.
``` {r function, echo=TRUE}
plot(cats$HeartWeight~cats$BodyWeight, main="Simple plot of heart weight \n as a functio
n of body weight", xlab="Body weight (kg)", ylab="Heart weight (g)")
fit<- lm(cats$HeartWeight~cats$BodyWeight)
abline(fit)
```